Solving \\ the \\ 15-Puzzle \\ \& \\ Rubik's Cube

### 2022-09-12 latex/Chapters/Rubiks.tex
trimming RC

In this second results section, I discuss my experiments on the Rubik's cube. I have run reasonably comprehensive experiments on the 2x2x2, throwing 6 of my solvers at the problem. As mentioned in \ref{sec:Puzzles}, the 2x2x2 \textbf{RC}, in spite of its deceivingly simple physical appearance, is actually not that much easier to solve (by hand) than the 3x3x3. For instance, it takes me usually just under 1 minute, wherras the 3x3x3 takes me only double that time, just under 2 minutes. I am obviously a beginner in Rubik's cube handsolving, but my personal ratio of time-to-solve and perceived difficulty is in line with that of experienced speedcubers who might take and average of, say, 2-3 seconds for the 2x2x2 and 10 seconds for the 3x3x3.
\\
Computing wise, the 2x2x2 was of decent size for my purpose, with its 88 million configurations. Let me start with a table summarizing which solvers I have attempted to run on the 2 dimensions I have attempted to experiment with:

\begin{center}
\begin{tabular}{l*{5}{c}r}
\hline
\textbf{solver}      & & \textbf{Rubik's} & \textbf{2x2x2} & \textbf{3x3x3} \\
\hline
BFS   &   &        &  x  &  x  \\
\hline
Kociemba   &   &      &  x  &  x  \\
\hline
A$^{*}$[Kociemba]  &   &  &  x  &  x  \\
\hline
A$^{*}$[DL[A$^{*}$[Kociemba]]]  &   &  &  x  &  x  \\
\hline
A$^{*}$[DRL]  &   &  &  x  &  \\
\hline
A$^{*}$[DQL]  &   &  &  x  &  \\
\hline
%MCTS[DQL][c=0]  &   &  & x &  \\
%\hline
%MCTS[DQL][c=69]  &   & & x &  \\
%\hline
\end{tabular}
\end{center}

@@

\Subsection{BFS}
BFS is the only one for which I used a step of 1 on the difficuly and stopped at 5 as it was clear it would not go any further. Given that the run time and expanded nodes are clearly going to increase linearly, and given where 5 scrambling from goal took this solver already, there was no point going any further!

\Subsection{Kociemba, A$^{*}$[Kociemba], A$^{*}$[DL[A$^{*}$[Kociemba]]]}
Next, let me discuss Kociemba (see \cite{HKociemba} and \cite{Kociemba}) and associated experiments. We obviously expect Kociemba to be by far the fastest solver since it is handcrafted with much specific knowledge about the Rubik's cube group theory and multi-stage solving which I discussed in \ref{Theory:RCMSS}. I found the 2x2x2 library I used to not be optimal however, despite claims to the contrary. I have not had the time to investigate why this was not the case, but empirically I found that for instance running A$^{*}$ with Kociemba's cost as a heuristic was resulting in higher quality (i.e. of lower cost) solutions, clearly showing the non-optimality of Kociemba to begin with.
\\
\noindent Similarly to the higher dimensions \textbf{SP} from the previous section, there was no hope for me to run a perfect solver on the 2x2x2 \textbf{RC}. For starter, I have no admissible heuristic to guarantee optimality! Even if I had found one such heuristic and managed to implement it, it is doubtful I could have solved all 88 millions configurations without considerable work and time to reduce the configurations via symmetry arguments. As a result, I used solutions from A$^{*}$[Kociemba] on (perfectly) randomly generated configurations to train my \textbf{DL} heuristic. Thinking of it, it is rather amusing since Kociemba is itself using a two-stage IDA$^{*}$ solver, so I end-up here with 3 levels of nested A$^{*}$ algorithms, namely A$^{*}$[DL[A$^{*}$[IDA$^{*}$[2-stage-subgroups]]]].
\\
Notice finally that, since I have no optimal solver for the \textbf{RC}, I have set Kociemba as my benchmark in the performance test for the purpose of computing the optimality score. This is why some (actually all) of the other solvers will show scores larger than 100\% (top-right graph of figure \ref{fig:222RCPerformance}).

\Subsection{Deep Reinforcement Learner \& Deep Q Learner}
\label{sec:S33DRL}
To get my \textbf{DRL} and \textbf{DQL} heuristics, I have run the DeepReinforcementLearner and DeepQLearner with the following main parameters:
\begin{itemize}
\item Update the target network every 1,000 epochs (\textit{update\_target\_network\_frequency=1000}) or when the MSE falls below 10 basis points of the maximum target cost-to-go (\textit{update\_target\_network\_threshold=1e-3})
\item Run for a maximum of 11,000 epochs (\textit{nb\_epochs=11000}) or when the maximum target so far has not been increasing by more than 1\% in the last 5,500 epochs (\textit{max\_target\_uptick=1e-2} and \textit{max\_target\_not\_increasing\_epochs\_pct=0.5})
\item At every update of the target network (\textit{training\_data\_every\_epoch=False}), generate 10,00 sequences of puzzles (\textit{nb\_sequences=10,000}), each comprised of a 15-scramble puzzle back to goal state, with all intermediary puzzles along the path (\textit{nb\_shuffles=15}).
\item Use a fully connected network (\textit{network\_type=fully\_connected\_net}) with three hidden layers of size 600, 300, 100 (\textit{layers\_description=(600,300,100)}) and use torch.optim.RMSProp as optimiser (\textit{optimiser=rms\_prop}) with an exponential scheduler (\textit{scheduler=exponential\_scheduler}) with gamma 0.999 (\textit{gamma\_scheduler=0.9999}) and \textit{learning\_rate=0.001}. The networks architecture is shown in figure \ref{fig:222RCnets}.
\end{itemize}
Graphs \ref{fig:222RCDRL} and \ref{fig:222RCDQL} show both learners' following metrics tracked over the epochs:
\begin{itemize}
\item learning rate, which decreases due to the scheduler, but gets reset at each network update.
\item max target. One interesting dynamic during learning is that due to the way targets are constructed using value-iteration update, and starting from an all 0 network, the max target only increases by about 1 at best at every network update. That is, the first network only learns how to differentiate goal from non-goal, then the second one learns how to differentiate goals, cost 1 and cost 2+ \dots. This is why I have introduced the \textit{update\_target\_network\_threshold} discussed above. It makes sure we do not waste time initially training the left-hand-side network when training is easy. \textit{max\_target\_uptick} and \textit{max\_target\_not\_increasing\_epochs\_pct} on the other hand make sure that we stop training altogether when the max target does not grow anymore. This is because we do not know a-priori what the God number of a given puzzle/dimension is. Let us assume for the sake of the argument that it is 15: that means that beyond 15 updates of the target network we might not get much benefit to additional training and value-iterating.
\item the loss (MSE for \textbf{DRL}, MSE + CrossEntropyLoss for \textbf{DQL})
\item the loss divided by the max target. The exit criterion \textit{update\_target\_network\_threshold} is based on that quantity, since a loss of e.g. 0.01 means little if the we do not compare it to the possible range that the cost-to-go can take.
\item The percentage of all the possible puzzles of that dimension \textit{seen} by the learner. This is an interesting quantity, as it tells us how well the solvers are able to generalise. From both graphs (\ref{fig:222RCDRL} \& \ref{fig:222RCDQL}) we can see that the learners saw less than 1\% of all puzzles. It is good to keep in mind however that \textit{seen} does not mean really much here, since the learning is \textit{unsupervised}. This is quite in contrast to the DeepLearner, whose traiing data consist of puzzles with their cost-to-go computed by a \textit{teacher} (optimal or other efficient solver).
\end{itemize}

\begin{figure}
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.6]{./Figures/222RCnets}}
  \caption[222RCnets]{2x2x2 \textbf{RC} architecture for \textbf{DRL} (left) \& \textbf{DQL} (right) heuristics training}
  \label{fig:222RCnets}
\end{figure}

\begin{figure}
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.5]{./Figures//222RCDRL}}
  \caption[222RCDRL]{\textbf{DRL} of the 2x2x2 \textbf{RC}}
  \label{fig:222RCDRL}
\end{figure}

\begin{figure}
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.5]{./Figures//222RCDQL}}
  \caption[222RCDQL]{\textbf{DQL} of the 2x2x2 \textbf{RC}}
  \label{fig:222RCDQL}
\end{figure}

@@

We summarize as usual the 2x2x2 \textbf{RC} performance tests by our usual metrics \ref{fig:222RCPerformance}. Overall, the \textbf{DL} and \textbf{DRL} managed to solve all cubes (though \textbf{DL} took over an houd and \textbf{DRL} up to 6 hours for the slowest cube they encountered. Both methods outperformed in terms of optimality score Kociemba and got on par with A$^{*}[Kociemba]$. Their max cost over all cubes encountered was 12 for up to 20 shuffles, and 13 for $\infty$ shuffling. Considering God's number for the 2x2x2 \textbf{RC} of 11, this is rather impressive!
\\
\\
An very interesting remark is that if we check the learning graph \ref{fig:222RCDRL} of the \textbf{DRL}, we can see it only ever assigned a cost-to-go (value function) of 7.6 at most. Despite that fact, it was still able to solve cubes requiring 13 moves. How is that possible? Notice that the loss of 0.2 towards which it converged at the end of its training is misleading, since it is not a loss versus the actual true cost-to-go, but rather versus that of the target network. My theory is that while the value the \textbf{DRL} assigned to hard-to-solve cubes is off (by at least 4 or 5), it is likely the case that the network gets the relative value of neighbouring cubes about right (given that the value-iteration teaches the solver that, it is not super surprising either), and this is sufficient to guide A$^{*}$ extremely well. If both the absolute value of the cost-to-go and the relative (among neighbouring cubes) was totally off, we would indeed not expect the heuristic to make A$^{*}$ any better than, say, \textbf{BFS} and we would therefore not expect it to be able to solve \textit{costly} cubes in 10-13 moves.
\\
\\
Finally, we can see that \textbf{DQL}, despite showing seemingly/visually cleaner and more monotonic loss convergence than \textbf{DRL} (see \ref{fig:222RCDQL} versus \ref{fig:222RCDRL}), struggled a lot more to solve difficult cubes (above 10 shuffles). Clearly this shows that my earlier findings on the 3x3 \textbf{SP} are likely not universal, and very much depend on the parameters set during training, as well as on the problem at hand.

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.55]{./Figures/222RCPerformance}}
  \caption[222RCPerformance]{Solvers' performance comparison 2x2x2 \textbf{RC}}
  \label{fig:222RCPerformance}
\end{figure}

@@

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.55]{./Figures/333RCPerformance}}
  \caption[333RCPerformance]{Solvers' performance comparison 3x3x3 \textbf{RC}}
  \label{fig:333RCPerformance}
\end{figure}

>>>>

I now discuss my experiments on the Rubik's cube. I have run reasonably comprehensive trial on the 2x2x2, throwing 6 of my solvers at the problem. As mentioned in \ref{sec:Puzzles}, the 2x2x2 \textbf{RC}, in spite of its deceivingly simple physical appearance, is actually not that much easier to solve (by hand) than the 3x3x3. For instance, it takes me usually just under one minute, versus two minutes for the 3x3x3. I am obviously no \textit{speedcuber}, but my time-to-solve and perceived difficulty ratios are in line with that of experienced players who might take 2-3 seconds for the 2x2x2 and 10 seconds for the 3x3x3. Computing wise, the 2x2x2 was of decent size for my purpose, with its 88 million configurations. Table \ref{tab:gridRC} summarizes which solvers I have attempted to run on the 2 dimensions:

\begin{table}[H]
\begin{center}
\begin{tabular}{l*{5}{c}r}
\hline
\textbf{solver}      & & \textbf{Rubik's} & \textbf{2x2x2} & \textbf{3x3x3} \\
\hline
BFS   &   &        &  x  &  x  \\
\hline
Kociemba   &   &      &  x  &  x  \\
\hline
A$^{*}$[Kociemba]  &   &  &  x  &  x  \\
\hline
A$^{*}$[DL[A$^{*}$[Kociemba]]]  &   &  &  x  &  x  \\
\hline
A$^{*}$[DRL]  &   &  &  x  &  \\
\hline
A$^{*}$[DQL]  &   &  &  x  &  \\
\hline
%MCTS[DQL][c=0]  &   &  & x &  \\
%\hline
%MCTS[DQL][c=69]  &   & & x &  \\
%\hline
\end{tabular}
\caption{\label{tab:gridRC} Solvers used vs \textbf{RC} dimension}
\end{center}
\end{table}

@@

\Subsection{BFS}
For \textbf{BFS} I used a step of 1 on the difficuly and stopped at 5 as it was clear it would not go much further. Its run time and expanded nodes are increasing linearly with difficulty, and it was already taking an average close to a minute for 5 shuffles from goal.

\Subsection{Kociemba, A$^{*}$[Kociemba], A$^{*}$[DL[A$^{*}$[Kociemba]]]}
I obviously expected Kociemba (see \cite{HKociemba} and \cite{Kociemba}) to be by far the fastest solver since it is handcrafted with much specific knowledge about the \textbf{RC} group theory and multi-stage solving (see section \ref{Theory:RCMSS}). Contrary to its claims, I found the 2x2x2 library to not be optimal, running A$^{*}$ with Kociemba's cost as a heuristic indeed resulted in shorter solutions. I did however set Kociemba as my benchmark in the \textit{performance\_test} since I had no admissible heuristic to guarantee optimality otherwise. The optimality scores (top-right graph of figure \ref{fig:222RCPerformance}) therefore do reach more than 100\%, as all other solvers found shorter solutions than Kociemba. I used the solutions of A$^{*}$[Kociemba] on perfectly randomly generated configurations to train the \textbf{DL} heuristic. Since Kociemba is itself using a two-stage IDA$^{*}$ solver, I end-up here with 3 levels of nested A$^{*}$ algorithms: A$^{*}$[DL[A$^{*}$[IDA$^{*}$[2-stage-subgroups]]]]!

\Subsection{Deep Reinforcement Learner \& Deep Q Learner}
\label{sec:S33DRL}
To get my \textbf{DRL} and \textbf{DQL} heuristics, I have run the DeepReinforcementLearner and DeepQLearner with the following main parameters:
\begin{itemize}
\item Update the target network every 1,000 epochs (\textit{update\_target\_network\_frequency}) or when $\frac{MSE}{max target}$ falls below 0.1\% (\textit{update\_target\_network\_threshold=1e-3})
\item Stop at \textit{nb\_epochs}=11,000 or if the max target has not increased more than 1\% (\textit{max\_target\_uptick}) in the last 5,500 epochs (\textit{max\_target\_not\_increasing\_epochs\_pct}=0.5)
\item At each target net update (\textit{training\_data\_every\_epoch}=False), generate \textit{nb\_sequences}=10,000 sequences of puzzles comprised of a 15-scramble puzzle along with its path to goal (\textit{nb\_shuffles}=15).
\item Fully connected network (\textit{network\_type=fully\_connected\_net}) with three hidden layers (\textit{layers\_description}=(600,300,100)), an RMSProp optimiser (\textit{optimiser=rms\_prop}), \textit{scheduler=exponential\_scheduler} with \textit{gamma\_scheduler}=0.9999 and \textit{learning\_rate}=0.1\%. The networks architecture are shown in figure \ref{fig:222RCnets}.
\end{itemize}

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.6]{./Figures/222RCnets}}
  \caption[222RCnets]{2x2x2 \textbf{RC} architecture for \textbf{DRL} (left) \& \textbf{DQL} (right) heuristics training}
  \label{fig:222RCnets}
\end{figure}

\noindent Graphs \ref{fig:222RCDRL} and \ref{fig:222RCDQL} show both learners' following metrics tracked over the epochs:
\begin{itemize}
\item learning rate: decreasing due to the scheduler and reset at each network update.
\item max target: an interesting dynamic during learning is that since targets are constructed using value-iteration update, starting from an all 0 network, the max target increases by at most about 1 at every network update, until we reach God's number. The first network only learns to differentiate goal from non-goal, the second one learns how to differentiate goals, cost 1 and cost 2+ \dots. This is why I introduced \textit{update\_target\_network\_threshold} discussed above, to makes sure we do not spend time initially training the left-hand-side network when it is easy. \textit{max\_target\_uptick} and \textit{max\_target\_not\_increasing\_epochs\_pct} make sure that we stop training when the max target does not grow anymore.
\item loss: MSE for \textbf{DRL}, MSE + CrossEntropyLoss for \textbf{DQL}
\item loss divided by max target: the exit criterion \textit{update\_target\_network\_threshold} is based on that quantity, since a loss of e.g. 1\% means little if the we do not compare it to the possible range of the cost-to-go.
\item The percentage of all the possible puzzles of that dimension \textit{seen} by the learner. This is an interesting quantity, as it tells us how well the solvers are able to generalise. From both graphs (\ref{fig:222RCDRL} \& \ref{fig:222RCDQL}) we can see that the learners saw less than 1\% of all puzzles. It is good to keep in mind however that \textit{seen} does not mean really much here, since the learning is \textit{unsupervised}. This is quite in contrast to the DeepLearner, whose traiing data consist of puzzles with their cost-to-go computed by a \textit{teacher} (optimal or other efficient solver).
\end{itemize}

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.4]{./Figures//222RCDRL}}
  \caption[222RCDRL]{\textbf{DRL} of the 2x2x2 \textbf{RC}}
  \label{fig:222RCDRL}
\end{figure}

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.4]{./Figures//222RCDQL}}
  \caption[222RCDQL]{\textbf{DQL} of the 2x2x2 \textbf{RC}}
  \label{fig:222RCDQL}
\end{figure}

@@

I summarize as usual the 2x2x2 \textbf{RC} performance tests with my usual metrics in figure \ref{fig:222RCPerformance}. Overall, the \textbf{DL} and \textbf{DRL} managed to solve all cubes (though \textbf{DL} took over an hour and \textbf{DRL} 6 hours for their slowest cube. Both methods outperformed Kociemba in terms of optimality-score and got on par with A$^{*}[Kociemba]$. Their max cost over all cubes encountered was 12 for up to 20 shuffles, and 13 for $\infty$ shuffling. Considering God's number for the 2x2x2 \textbf{RC} of 11, this is rather impressive!
\\
\\
Quite remarkably we can see (second graph of \ref{fig:222RCDRL}) that the \textbf{DRL} heuristic only ever assigned a cost-to-go of 7.6 at most. Despite that, it was still able to solve cubes requiring up to 13 moves (likely to have been among the hardest 2x2x2 cubes). How is that possible? Notice that the loss of 0.2 towards which the network converged at the end of its training is misleading, since this is in-sample loss versus the target network. My theory is that while the value the \textbf{DRL} assigned to hard-to-solve cubes can be off (by at least 4 or 5), it is likely that the network gets the relative value of neighbouring cubes right (not that surprising given the value-iteration), and this is sufficient to guide A$^{*}$ extremely well. If both the absolute value of the cost-to-go and the relative (among neighbouring cubes) was totally off, we would indeed not expect the heuristic to make A$^{*}$ any better than, say, \textbf{BFS} and we would therefore not expect it to be able to solve \textit{costly} cubes requiring 10+ moves.
\\
\\
Finally, we can see that \textbf{DQL}, despite showing visually cleaner and more monotonic loss convergence than \textbf{DRL} (see \ref{fig:222RCDQL} versus \ref{fig:222RCDRL}), struggled a lot more to solve difficult cubes (above 10 shuffles). This shows that my earlier findings on the 3x3 \textbf{SP} are likely not universal, and very much depend on the parameters set during training, as well as on the problem at hand.

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.52]{./Figures/222RCPerformance}}
  \caption[222RCPerformance]{Solvers' performance comparison 2x2x2 \textbf{RC}}
  \label{fig:222RCPerformance}
\end{figure}

@@

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.52]{./Figures/333RCPerformance}}
  \caption[333RCPerformance]{Solvers' performance comparison 3x3x3 \textbf{RC}}
  \label{fig:333RCPerformance}
\end{figure}

### 2022-09-12 latex/Chapters/SlidingPuzzle.tex
trimming RC

I now discuss results on the \textbf{SP}, going from \textit{small} dimensions (which I can reasonably \textit{fully} solve given my computing resources, the finite time I have to complete this project, and the choice of programming language), then I move to studying in detail the 3x3 \textbf{SP}, and finally move to higher 3x4 and 4x4 dimensions. Table \ref{tab:gridSP} shows which solvers I have run on each dimension. The reader is referred to sections \ref{sec:Chapter1} and \ref{sec:Implementation} for details on each solver, and to the appendix section \ref{sec:Examples} for code examples.

>>>>

I now discuss results on the \textbf{SP}, going from \textit{small} dimensions (which I can reasonably \textit{fully} solve given my computing resources, the finite time I have to complete this project, and the choice of programming language), then I move to studying in detail the 3x3 \textbf{SP}, and finally move to higher 3x4 and 4x4 dimensions. Table \ref{tab:gridSP} shows which solvers I have run on each dimension. The reader is referred to the theory (\ref{Chapter1}) and implementation (\ref{sec:Implementation}) sections for details on each solver, and to the appendix section \ref{sec:Examples} for code examples.
